Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest

### 2019-06-19 manuscript/03-methods.Rmd
Added Helcoski et al 2019 source

## Data Collection
Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013). In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016]. They were sanded, measured, and standardized in the same process as seen in Helcoski et al 2019.

@@

To determine focus drought years, we used the pointRes package in R (version 3.6.0) to determine resilience metrics. Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the past 5 years. The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA. For ease of research, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.

Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from Jennings 1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed). Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species’ cores. In total, 14 species had cores that were considered complete and usable, as seen in Table 1.

Geographic info

@@

Growth

As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems. Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,

diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))

First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species. Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness in from the regression equations.

Sapwood ratio also used the bark thickness data, and was also calculated for 2008, where

sapwood_area = total wood area (without bark) - heartwood area, and
sapwood_Ratio = sapwood area / total wood area (without bark)

Once sapwood area for each tree was calculated, log-log regression equations were determined on a per-species basis, which were then used

Height

Height data for trees was obtained from clinometers and rangefinders, from 2012 to 2019. Log=log regression equations were developed for species based on this height data, then heights of the study trees were calculated using these equations. For species that didn't have enough height measurements, heights were calculated from equations derived from all study-focus species.

Height sources include [@stovall_assessing_2018], [@stovall_terrestrial_2018], [@andersonteixeira_ctfs-forestgeo:_2015],

Traits
Turgor loss point (TLP) was determined per species from leaf trait studies carried out at sCBI in 2018. TLP was determined using __________________ on ### of individuals from each species (https://github.com/EcoClimLab/HydraulicTraits/blob/master/data/SCBI/processed_trait_data/SCBI_LeafTraits_measurements_plus_intermediary.csv).  Ring porosity was qualified from Anderson-Teixeira et al 2015.

>>>>

## Data Collection
Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013). In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016]. They were sanded, measured, and standardized in the same process as seen in [@helcoski_growing_2019].

@@

To determine focus drought years, we used the pointRes package in R (version 3.5.2) to determine resilience metrics. Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the past 5 years. The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA. For ease of research, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.

Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species’ cores. In total, 14 species had cores that were considered complete and usable, as seen in Table 1.

|**Species**|**n_cores.canopy**|**n_cores.subcanopy**|**included**
|----------|-------------|---------------|-----------|
*Carya cordiformis*|2|11| sub-canopy only
*Carya glabra*|9|22|yes
*Carya ovalis*|9|14|yes
*Carya tomentosa*|0|13|sub-canopy only
*Fagus grandifolia*|7|73|yes
*Fraxinus americana*|20|42|yes
*Fraxinus nigra*|1|11| sub-canopy only
*Juglans nigra*|22|9|yes
*Liriodendron tulipifera*|38|60|yes
*Pinus strobus*|14|21|no
*Quercus alba*|38|23|yes
*Quercus prinus*|27|32|yes
*Quercus rubra*|43|26|yes
*Quercus velutina*|54|23|yes

``` {r sample, echo=FALSE}
#numbers for the table above come from this table
# library(RCurl)
# library(dplyr)
# species <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/master/data/core_list_for_neil.csv?token=AJNRBELCYZY576ARLE7BXNK5BEF3O"))
# 
# test <- species %>%
#   group_by(sp, crown.position) %>%
#   summarize(count=n())

```

### Biophysical traits
Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed. For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).

@@

Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. The associated errors for using either method were acknowledged [@larjavaara_measuring_2013]. Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees. For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.

Equations required the DBH in each individual drought year. As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems. Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,

`diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))`

First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species. Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.

### Leaf Hydraulic Traits
Hydraulic traits were collected from SCBI and are summarized in Table 1.

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

traits_table <- data.frame(
  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Chlorophyll", "Wood density", "TLP", "P50", "P80"),
  "Unit" = c("", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"),
  "Source" = c("[@andersonteixeira_ctfs-forestgeo:_2015]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]"))

kable(traits_table) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

```
